Evolutionary Design of Particles for Collectively Intelligent Problem Solving

### 2014-05-08 docs/paper.tex

Self-organizing "flocks" of agents that display emergent coordinated behavior can be implemented using only local interaction rules. When agents are also given memory and a goal-based Finite State Machine (FSM) controller, the resulting team is able to solve a locate-and-collect task. We show that Evolutionary Computing techiques can be applied to the parameters that comprise the FSM states. Comparing the resulting FSMs to similar controllers described in the previous literature, we find that the evolved FSMs perform as well as or better than those designed by human  observation. We close by suggesting methods for applying evolutionary techniques to the structure as well as to the numerical parameters of the controller, perhaps leading to the emergence of novel behavior.

@@

Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems \cite{reynolds1987flocks}, wherein each agent's motion is determined solely by local interactions with other agents, there has been interest in extending such systems to solve more general problems. The self-organizing characteristic of these systems, whereby coordinated behavior emerges without a central control mechanism, suggests that even relatively simple decentralized systems might yield novel approaches to problems in both the computational and physical domains.

By abstracting the idea of "flocks" to $n$-dimensional spaces, for example, "particle swarms" \cite{kennedy1995particle,clerc2002particle} or "cultural algorithms" \cite{chung1996testbed} have been used for numerical optimization. In the physical domain, swarm behavior have been used multirobot team movement control \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, or during deployment of mobile robots or sensors. In \cite{cheng2009distributed}, flocking behavior was used to maximize coverage during such deployments while still maintaining collective behavior. Other tasks being explored include urban pursuit \cite{winder2004using} and robot herding of animals \cite{vaughan1998robot}.

Many approaches augment the agents with such features as working memory \cite{winder2012role,hu2003particle}. Rodriguez \cite{rodriguez2004extending} added both memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal. He was able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.

Attempts have been made to formalize the design of such cooperative multi-agent systems \cite{mataric1993designing,capera2003amas}. In this paper, however, we turn to evolutionary computation techniques. Genetic Programming (GP) \cite{koza1992genetic}, Evolutionary Strategies \cite{rechenberg1989evolution}, and Genetic Algorithms \cite{goldberg1988genetic} have demonstrated the potential of computational design techniques for machine learning, optimization, and even engineering. Finite state machines have been evolved to solve problems such as automatic target detection \cite{benson2000evolving}. Evolving the FSM described by Rodriguez, therefore, seems like a natural extension of multi-agent problem-solving. This is not to be confused with evolutionary techniques that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}.

In this paper, we have attempted to extend Rodriguez's findings in \cite{rodriguez2004extending} by applying evolutionary techniques specifically to the agent control mechanisms. Rodriguez determined the structure of the FSM and the parameters defining each state through inspection and empirical techniques. We attempt to determine the optimal parameters by an evolutionary process that runs a population of randomly-generated FSM configurations through a simulated problem and then evolves the population through fitness-based selection, recombination, and mutation genetic operators.

@@

@@

Avoidance: a vector pointing away from an enemy agent in the neighborhood, with magnitude $v_{max}$ when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is r. Unlike all the other components, this can be applied multiple times each timestep--once for every enemy in the neighborhood.

@@

At the beginning of each timestep, therefore, each agent is in a particular state. For each velocity component in its current state, it determines the number of friendly and enemy agents in the respective neighborhood and calculates the direction and magnitude of the component. These components are multiplied by the respective weight, added to the agent's velocity from the previous timestep, and the resulting velocity is subjected to the $v_{max}$ constraint. The agent's position is updated by adding the velocity to the agent's position in the previous timestep. The agent then determines whether it should change its state based on the predefined transition rules and its current environment.

@@

After running each of the 50 configurations, we subjected the population to a round of evolution. The most fit individual from the current generation was always carried into the next generation, with the other 49 selected via tournaments of size 3. Each of the 49 was then subjected to mutation, with an independent 20\% chance of mutating each weight, each radius, and each alpha of every component. If selected for mutation, the trait was changed by a random value uniformly-distributed between -0.2 and 0.2 for weights, -20 and 20 for radii, and -20 and 20 for alphas. Additionally, the home guarding and deposit guarding threshold each had a 20\% of changing by plus or minus 1.

>>>>

Emergent, self-organizing flocks of particles can be implemented using only simple, local movement behavior rules. When particles are also improved with a limited working memory and goals, the local interactions of these top-down controllers leads to emergent global problem solving such as search and retrieve. We will show in this paper that the rules for both movement and goals can be evolved using Evolutionary Computing techiques on the dynamics parameters, as well as on the finite state machines that control behavior. The evolved particles perform as well as or better then those designed by humans.

@@

Ever since Reynolds demonstrated that flocks of birds could be simulated as multi-agent systems \cite{reynolds1987flocks}, there has been interest in improving flocks, schools, and swarms to higher levels of collective intelligence. The relatively simple nature of each agent; its motion or goals are determined solely by local interactions; leads to emergent global behavior. Decentralized control would mean optimal and economic solutions to problems in both algorithmic and physical domains. Furthermore, there has been interest in extending such systems to solve more general problems than traditional optimization \cite{kennedy1995particle}, modeling complex systems \cite{ottino2003complex,clerc2002particle} or cultural algorithms \cite{chung1996testbed}.

This extension has primarily focused on the physical domain to improve multirobot team movement control as in \cite{balch1998behavior,ccelikkanat2010steering,hodgins1994robot}, however higher order problem solving is currently being explored. One large area of research is the deployment of mobile robots or sensors for area coverage, either to map terrain, or to establish mesh communication links. In \cite{cheng2009distributed}, flocking behavior was used to maximize coverage, while still maintaining collective behavior. Other tasks being explored include urban pursuit \cite{winder2004using} and robot herding of animals \cite{vaughan1998robot}.

Many approaches have improved the simple local dynamics by adding minimal, cost effective top down components that also behave locally to produce emergent behavior. One very successful addition to flocks for problem solving is the addition of working memory \cite{winder2012role,hu2003particle}. Building on flocks with memory improvements, Rodriguez in \cite{rodriguez2004extending} added goal-driven intelligence in the form of a finite state machine (FSM) that switched the agent between different sets of movement behaviors according to its local environment and current goal. He was able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.

These exciting results seem to indicate that in the future difficult environments might be economically explored or exploited by flocking robotic teams exhibiting local behavior for collective intelligence. For example in disaster areas such as fires or earthquakes, teams of flocking robots may be able to more effectively search and rescue victims. Landscapes such as other planetary bodies, the deep sea, or volcanos may similarly be accessible by such teams. However, in order to work in varied terrains and tasks specific swarm systems must be developed by humans to design both the bottom up movement behaviors and the top down controllers. Such design tasks have even led to formal design theory such as \cite{mataric1993designing,capera2003amas}. Although this theory might be applicable to a wide array of problems, a more flexibile approach could enable even more adaptive systems.

Familiar evolutionary computation techniques such as Genetic Programming (GP) \cite{koza1992genetic}, Evolutionary Strategies \cite{rechenberg1989evolution}, and Genetic Algorithms \cite{goldberg1988genetic} have already shown proof of the potential of computational design techniques for machine learning, optimization, and even engineering. Finite state machines have already undergone evolutionary processes to produce better results for particular tasks like automatic target detection \cite{benson2000evolving}. Evolving movement behaviors would be the equivalent of other real value optimization problems. Using these techniques it seems that evolving the individual particles in a swarm would be an effective technique of natural design and adaptation that mimics nature.

This is not to be confused with evolutionary technqiues that leverage swarms as in \cite{wei2002swarm,miranda2005evolutionary}. In this paper, we have attempted to extend Rodriguez's findings in \cite{rodriguez2004extending} by applying evolutionary techniques specifically to the agent control mechanisms. Rodriguez determined the structure of the FSM and the parameters defining each state through inspection and empirical techniques. We attempt to determine the optimal parameters by an evolutionary process that runs a population of randomly-generated FSM configurations through a simulated problem and then evolves the population through fitness-based selection, recombination, and mutation genetic operators.

@@

%TODO: Discuss maximum velocity and other simulation parameters
%TODO: Discuss how the velocity is computed given the velocity components
%TODO: Dicsuss the finite state machine and how it is composed of behaviors

@@

Avoidance: a vector pointing away from an enemy agent in the neighborhood, with magnitude Vmax when distance to the enemy is zero and decreasing linearly with distance until it is zero when the distance is r. Unlike all the other components, this can be applied multiple times each timestep--once for every enemy in the neighborhood.

@@

At the beginning of each timestep, therefore, each agent is in a particular state. For each velocity component in its current state, it determines the number of friendly and enemy agents in the respective neighborhood and calculates the direction and magnitude of the component. These components are multiplied by the respective weight, added to the agent's velocity from the previous timestep, and the resulting velocity is subjected to the Vmax constraint. The agent's position is updated by adding the velocity to the agent's position in the previous timestep. The agent then determines whether it should change its state based on the predefined transition rules and its current environment.

@@

The evolution of a finite state machine for movement control was achieved through the application of a simple set of genetic operators to the weights and parameters of the machine. The genotype of a movement controller was represented as two integers, one for each guarding threshold, and a vector of real-valued numbers consisting of the alpha parameter, radius, and weight given to each component of each compound movement behavior.

The set of states in each controller was fixed to be the same set of compound behaviors as those found in \cite{rodriguez2004extending}, but with parameters derived through evolutionary programming. This allowed for easy comparison with the authoritative, human-derived solution. In addition, the transitions between states were kept the same except for the guarding thresholds, which we evolved. New individuals were introduced through the application of mutation and recombination operators.

The mutation operator first selected which parameters to modify independently. We used a 20\% chance for mutating each weight, radius, alpha, and threshold. Each selected threshold had an equal chance of increasing by 1 and decreasing by 1. Each selected real-valued parameter was modified through the addition of a uniformly distributed random number with mean 0. We chose to mutate each weight by -0.2 to 0.2 and each alpha and radius by -20 to 20.

The recombination operator ... %TODO talk about recombination

A round of evolution combined the mutation and recombination operators to produce the next generation. After evaluating every individual in the current generation, we applied elitism, tournament selection, mutation, and recombination to complete a round of evolution. For elitism, the single best individual of the previous generation was always carried into the next generation without modification. The other 49 individuals were selected via tournaments of size 3. Each of the 49 was then subjected to the mutation operator, with its parameters being mutated independently.
